\documentclass{article}

% preamble.tex from CSCI 693 / CCSC template
\input{preamble}

\addbibresource{paper.bib}

\graphicspath{{figures/}}

\title{Rust vs.\ C++ for CPU and GPU Performance: A Controlled Empirical Benchmark Study%
  \footnote{CSCI 693 Research Methods in Computer Science, Spring 2026}}

\author{
  Aditya Raut\affmark[1], Shelley Wong\affmark[2]\\
  \affmark[1]Computer Information Systems\\
  \affmark[2]Computer Science Department\\
  California State University, Chico\\
  Chico, CA 95929\\
  \email{\{araut1,swong26\}@csuchico.edu}
}

\begin{document}
\maketitle

% ─────────────────────────────────────────────────────────────
\begin{abstract}
This paper presents a controlled empirical study comparing the runtime
performance of Rust and C++ across CPU computational benchmarks and
CUDA-accelerated GPU workloads.
Benchmarks cover sorting, search, hash tables, dense matrix multiplication,
fast Fourier transform (FFT), $n$\nobreakdash-body simulation, and
reduction kernels.
All experiments were executed on a dedicated workstation equipped with an
Intel Core Ultra~7 processor and an NVIDIA RTX~5060~Ti GPU running
Ubuntu~24.04 and CUDA~12.4.
Statistical analysis applies Welch's $t$\nobreakdash-test, Cohen's~$d$
effect size, and 95\,\% confidence intervals to 5--30 timed trials per
benchmark.
Rust outperforms C++ on several CPU tasks---notably search, sorting, and
integer hash tables---while C++ retains advantages on FFT, naive matrix
multiplication, and one parallel-reduction configuration.
For the GPU kernels, Rust achieves significantly lower kernel times for
matrix multiplication ($p=0.017$, $d=-2.49$) and reduction
($p=0.001$, $d=-5.51$), while the softmax comparison yields no
statistically significant difference ($p=0.778$).
Safety-oriented metadata confirm zero \texttt{unsafe} blocks in the measured
Rust paths, and Rust requires fewer lines of code in aggregate.
The results are consistent with the hypothesis that Rust can be
performance-competitive across both CPU and GPU workloads without sacrificing
safety guarantees.
\end{abstract}

% ─────────────────────────────────────────────────────────────
\section{Introduction}

\subsection{Context and Background}

High-performance computing (HPC) continues to rely heavily on low-level
programming languages that afford explicit control over memory layout,
vectorisation, and hardware access~\cite{dongarra2011international}.
C++ remains the dominant implementation language in many
performance-critical CPU and GPU systems because of its mature compiler
ecosystem (Clang/GCC/MSVC), direct CUDA integration, and an extensive body
of optimised libraries~\cite{stroustrup2007evolving}.
Rust, however, has emerged as a strong alternative for systems programming by
combining ahead-of-time compilation with a memory-safe ownership model and
first-class concurrency guarantees~\cite{jung2017rustbelt}.

\subsection{Problem Statement and Research Questions}

The core question for practitioners is whether Rust can offer comparable
runtime performance to C++ while reducing safety risks and improving
maintainability.
This study operationalises that question through three hypotheses:

\begin{enumerate}[noitemsep]
  \item \textbf{H1 (CPU parity):} Rust runtimes on CPU workloads are within
    $\pm$15\,\% of equivalent C++ runtimes on average.
  \item \textbf{H2 (GPU parity):} Rust CUDA-backed GPU kernel times are
    within 5\,\% of equivalent C++ CUDA kernel times on average.
  \item \textbf{H3 (safety cost):} Competitive Rust performance is achieved
    without the use of \texttt{unsafe} blocks, indicating that safety
    guarantees carry no explicit opt-out cost in this benchmark set.
\end{enumerate}


\subsection{Significance of the Study}

Prior work suggests that Rust can approach C++ performance on CPU workloads
and may be viable for GPU-adjacent development when paired with CUDA
back-ends or driver bindings~\cite{bychkov2022rust, ivanov2022rust,
martins2025npb}.
However, no prior study has evaluated Rust and C++ together on both CPU and
GPU tasks using a single, unified benchmark project with end-to-end
statistical rigour and a consistent hardware environment~\cite{georges2007statistically}.
This study fills that gap by providing a controlled, reproducible comparison
tied to publicly committed artifacts.

\subsection{Artifact-Based Methodology and Scope}

All benchmarks in this study were executed on a single, dedicated workstation
(hardware specifications in Section~\ref{sec:methodology}) and the raw timing
outputs were committed to a public repository alongside the source code,
analysis scripts, and derived summary tables.%
\footnote{\url{https://github.com/adityaanilraut/Rust-vs-Cpp-for-CPU-and-GPU-Performance}}
This paper reports findings drawn from those committed artifacts rather than
rerunning benchmarks at manuscript-preparation time.
This choice is deliberate: it ties every numerical claim to a specific,
documented hardware configuration and software stack, and ensures
bit-for-bit reproducibility with the archived data.
Repeating the benchmarks on different hardware would introduce
platform-specific variability and make it harder to compare results against
the committed dataset.
Section~\ref{sec:methodology} documents all conditions under which the
original runs were performed, and the repository provides the raw data and
scripts needed to reproduce the analysis on the same or equivalent hardware.

% ─────────────────────────────────────────────────────────────
\section{Literature Review}

\subsection{Overview of Existing Research}

Research on Rust's performance relative to C++ has grown substantially since
Rust's 1.0 release in 2015.
Early studies focused on validating whether the ownership model and
borrow-checker impose runtime costs.
Zhang~et~al.\ characterise Rust's runtime behaviour across a broad set of
microbenchmarks and report that the compiler generates competitive code on
most workloads~\cite{zhang2022rust}.
Ivanov compares Rust with C++ on common numerical and data-structure routines
and finds that Rust can match or exceed C++ performance on many optimised CPU
tasks, with occasional overheads on workloads that rely heavily on
compiler-specific auto-vectorisation~\cite{ivanov2022rust}.

The conversation has since extended to HPC-scale benchmarks.
Martins~et~al.\ adapt the NAS Parallel Benchmarks to Rust and show that the
performance gap between Rust and established HPC toolchains (primarily C and
Fortran) is continuing to narrow, though some collective-communication
workloads retain non-trivial overheads~\cite{martins2025npb}.
The Stack Overflow Developer Survey consistently ranks Rust as the
most-admired language while noting that its adoption in production HPC
contexts remains limited relative to C++~\cite{stackoverflowSurvey2024},
which motivates studies like this one that quantify the actual performance
cost of that adoption.

On the GPU programming side, Bychkov and Nikolskiy examine Rust as a
language for GPU workloads and emphasise that fair comparisons require
equivalent kernel implementations and consistent memory-transfer
accounting~\cite{bychkov2022rust}.
K\"{o}pcke~et~al.\ introduce Descend, a safe GPU systems language inspired
in part by Rust's type system, reflecting broader research interest in
bringing safety guarantees to GPU programming~\cite{kopcke2023descend}.
Veytsman~et~al.\ provide an applied case study in computational physics,
arguing that Rust can eliminate classes of memory-related defects while
preserving execution performance~\cite{veytsman2024rewrite}.

\subsection{Critical Evaluation}

Taken together, the existing literature establishes three points of
consensus: (1)~Rust produces competitive CPU binaries on most workloads,
(2)~the borrow-checker and ownership model incur negligible runtime overhead
for well-written code, and (3)~GPU-oriented Rust work is nascent but
promising.
However, most studies evaluate either CPU or GPU performance---not both
together in the same controlled environment---and few apply full statistical
rigour (confidence intervals, effect sizes, and variance checks) to their
timing comparisons.
Bychkov and Nikolskiy's caution about equivalent kernels is methodologically
important but rarely operationalised in published comparisons.
The study by Fulton~et~al.\ further demonstrates that adopting Rust carries
real developer-experience benefits alongside safety gains, but these are
measured qualitatively rather than empirically~\cite{fulton2021benefits}.

\subsection{Research Gap and Rationale}

The gap this study addresses is the absence of a unified, hardware-specific,
statistically rigorous benchmark across both CPU and GPU workloads using
identical source repositories and analysis pipelines for Rust and C++.
Existing work either (a)~focuses only on CPU, (b)~uses informal timing
without confidence intervals or effect sizes, or (c)~evaluates GPU
performance in isolation from a broader CPU context.
This study contributes a controlled comparison that covers both domains,
applies Welch's $t$-tests and Cohen's $d$ to quantify statistical
significance and practical effect size, and also measures safety and
code-complexity proxies alongside performance.

% ─────────────────────────────────────────────────────────────
\section{Methodology}
\label{sec:methodology}

\subsection{Research Design}

This is a controlled empirical benchmark study.
Rust and C++ implementations of the same algorithms were compiled at the
highest optimisation level on the same machine, run against identical
input datasets, and timed over multiple independent trials.
Statistical tests were then applied to the resulting distributions to
evaluate each hypothesis.

\subsection{Hardware and Software Environment}

\begin{itemize}[noitemsep]
  \item \textbf{CPU:} Intel Core Ultra~7 (Meteor Lake, efficiency-core + performance-core hybrid)
  \item \textbf{GPU:} NVIDIA GeForce RTX~5060~Ti (Blackwell architecture)
  \item \textbf{Operating System:} Ubuntu 24.04 LTS (64-bit)
  \item \textbf{CUDA:} Version~12.4, \texttt{nvcc} V12.4.131
    (built Thu Mar~28 02:18:24~PDT~2024)
  \item \textbf{C++ compiler:} GCC via CMake Release build
    (\texttt{-O2 -DNDEBUG}); CUDA kernels compiled with \texttt{nvcc -O2}
  \item \textbf{Rust compiler:} \texttt{rustc} stable channel,
    \texttt{cargo build -{}-release} (\texttt{opt-level = 3})
  \item \textbf{CPU benchmark harness:} Google Benchmark (C++) and the
    standard-library timing approach (Rust); inputs defined by
    \texttt{generate\_datasets.py}
\end{itemize}

All C++ and Rust source files, build scripts, dataset generation utilities,
and analysis scripts are included in the public repository.

\subsection{Benchmark Configuration and Fairness}

\textbf{CPU benchmarks.}
Each CPU benchmark was executed under Google Benchmark's auto-iteration
mode, which determines the number of repetitions needed to achieve a stable
mean.
The timing metric is wall-clock time per operation, not CPU time, to
capture scheduler effects consistently across both implementations.

\textbf{GPU benchmarks.}
Each GPU kernel was run for five independent trials using separately
launched CUDA contexts.
Timing covers kernel execution only; host-to-device (H2D) and
device-to-host (D2H) transfer times are recorded separately and excluded
from the primary kernel-time comparison (though they are available in the
raw CSVs).
This separation follows the guidance of Bychkov and Nikolskiy~\cite{bychkov2022rust}
that transfer costs should not be conflated with kernel throughput in
language comparisons.

\textbf{Fairness measures.}
Both language implementations use algorithmically equivalent code for each
benchmark.
The same binary input datasets, generated by a single shared Python script,
are used for all C++ and Rust runs.
Compiler optimisation flags are set to the highest standard release level
for each ecosystem (\texttt{-O2}/\texttt{-O3} for C++, \texttt{opt-level = 3}
for Rust).
No language-specific SIMD intrinsics or assembly stubs were introduced.

\subsection{Statistical Analysis}

For each (benchmark, input-size) pair, the following statistics are computed
by \texttt{scripts/analyze\_results.py}:

\begin{itemize}[noitemsep]
  \item Mean and standard deviation across all timed trials
  \item 95\,\% confidence interval via the Student $t$-distribution
  \item Welch's two-sample $t$-test (does not assume equal variance)
  \item Cohen's $d$ effect size (pooled standard deviation formula)
  \item Coefficient of variation (CV\,\%) with a high-variance flag at
    CV\,$>$\,5\,\%
\end{itemize}

Percentage overhead is defined as
$(\mu_\text{Rust} - \mu_\text{C++})/\mu_\text{C++} \times 100$,
so negative values indicate Rust is faster.

\subsection{Lines-of-Code and Safety Metrics}

Lines of code (LOC) are counted by \texttt{scripts/safety\_analysis.py}
using the following rules: a line is counted if and only if it is
non-blank and does not begin with a single-line comment character
(\texttt{//} for C++ and Rust) or a block-comment delimiter
(\texttt{/*} or \texttt{*}).
The \texttt{build/} and \texttt{target/} directories are excluded.
The tool is applied to \texttt{.cpp}, \texttt{.h}, and \texttt{.hpp}
files for the C++ components and to \texttt{.rs} files for the Rust
components.

Rust \texttt{unsafe} blocks are detected by scanning every
\texttt{.rs} file (excluding \texttt{target/}) for lines matching the
regular expression \verb|\bunsafe\b|.
A ``recorded unsafe block'' therefore means a file-level occurrence of
the keyword \texttt{unsafe} in the measured source paths; it does not
include transitive \texttt{unsafe} inside dependencies.
The phrase ``measured paths'' refers specifically to the source files
under \texttt{cpu/rust/} and \texttt{gpu/rust/} in the repository.

\subsection{Limitations}

Several limitations affect how broadly these results can be generalised.
First, the GPU benchmark set contains only five trials per kernel, which
is smaller than the 30+ trials typically used for CPU measurements;
this produces wider confidence intervals and higher CV\,\% values for
the GPU results.
Second, the benchmark suite covers a focused selection of kernels
(sorting, search, linear algebra, FFT, reduction) and does not
represent all HPC or AI workloads.
Third, all results are tied to a single hardware platform; results may
differ on CPUs with different microarchitectures or GPUs of different
generations.
Fourth, LOC and \texttt{unsafe} counts are proxies for safety and
maintainability---they are necessary but not sufficient evidence for
claims about developer experience.

% ─────────────────────────────────────────────────────────────
\section{Results}

\subsection{CPU Benchmark Outcomes}

Table~\ref{tab:cpu_results} presents the complete CPU results drawn from
the committed summary table.
The table covers all 15 CPU benchmarks with mean runtimes, percentage
overhead, $p$\nobreakdash-values (Welch's $t$-test), and Cohen's $d$.
Negative overhead indicates that Rust is faster than C++;
positive overhead indicates that C++ is faster.

\begin{table}[htbp]
\caption{CPU performance results: C++ vs.\ Rust (all times in ms,
         mean $\pm$ std).}
\label{tab:cpu_results}
\centering
\begin{tabular}{llrrrr}
\toprule
Benchmark & Size & C++ (ms) & Rust (ms) & Overhead & $p$-value \\
\midrule
binary\_search           & 1\,000\,000 & 7.20 $\pm$ 0.03 & 2.28 $\pm$ 0.00 & $-$68.3\,\% & $<$0.001$^{***}$ \\
lower\_bound             & 1\,000\,000 & 7.26 $\pm$ 0.01 & 2.10 $\pm$ 0.01 & $-$71.1\,\% & $<$0.001$^{***}$ \\
hashmap\_int\_insert     & 100\,000   & 3.78 $\pm$ 0.00 & 1.08 $\pm$ 0.00 & $-$71.3\,\% & $<$0.001$^{***}$ \\
hashmap\_int\_lookup     & 100\,000   & 1.20 $\pm$ 0.00 & 0.97 $\pm$ 0.00 & $-$19.8\,\% & $<$0.001$^{***}$ \\
hashmap\_string\_insert  & 100\,000   & 8.39 $\pm$ 0.07 & 6.16 $\pm$ 0.02 & $-$26.6\,\% & $<$0.001$^{***}$ \\
quicksort\_std\_sort     & 1\,000\,000 & 49.70 $\pm$ 0.04 & 9.62 $\pm$ 0.02 & $-$80.6\,\% & $<$0.001$^{***}$ \\
quicksort\_manual        & 1\,000\,000 & 54.47 $\pm$ 0.13 & 44.84 $\pm$ 0.04 & $-$17.7\,\% & $<$0.001$^{***}$ \\
mergesort\_recursive     & 1\,000\,000 & 63.30 $\pm$ 0.11 & 53.55 $\pm$ 0.29 & $-$15.4\,\% & $<$0.001$^{***}$ \\
nbody\_single\_step      & 1\,024     & 2.01 $\pm$ 0.00 & 2.02 $\pm$ 0.00 & $+$0.6\,\%  & 0.007$^{**}$ \\
nbody\_5\_timesteps      & 1\,024     & 10.04 $\pm$ 0.00 & 10.06 $\pm$ 0.03 & $+$0.3\,\% & 0.033$^{*}$ \\
parallel\_reduction\_seq & 1\,000\,000 & 0.39 $\pm$ 0.00 & 0.38 $\pm$ 0.00 & $-$1.9\,\%  & $<$0.001$^{***}$ \\
matmul\_naive            & 512        & 14.23 $\pm$ 0.06 & 18.57 $\pm$ 0.02 & $+$30.5\,\% & $<$0.001$^{***}$ \\
fft\_forward             & 65\,536    & 0.93 $\pm$ 0.00 & 1.22 $\pm$ 0.00 & $+$30.9\,\%  & $<$0.001$^{***}$ \\
fft\_roundtrip           & 65\,536    & 1.93 $\pm$ 0.00 & 2.40 $\pm$ 0.00 & $+$24.4\,\%  & $<$0.001$^{***}$ \\
parallel\_reduction\_rayon & 1\,000\,000 & 0.03 $\pm$ 0.00 & 0.10 $\pm$ 0.01 & $+$192.2\,\% & $<$0.001$^{***}$ \\
\bottomrule
\end{tabular}
{\small $^{*}p<0.05$, $^{**}p<0.01$, $^{***}p<0.001$;
 CV\,\% flags omitted for brevity---high-variance entries:
 \texttt{parallel\_reduction\_rayon} (C++ CV\,3.3\,\%, Rust CV\,12.1\,\%).}
\end{table}

Rust shows statistically significant advantages (all $p\,<\,0.001$) on
search, hash-table, and sorting benchmarks.
The standard-library sort advantage ($-$80.6\,\%) likely reflects
algorithmic differences between C++ \texttt{std::sort} (introsort) and
Rust's pattern-defeating quicksort (pdqsort), rather than a language-level
performance difference.
For matrix multiplication and FFT, C++ retains significant advantages;
those benchmarks are sensitive to auto-vectorisation quality and
library-level optimisation that may favour the C++ compiler on this
microarchitecture.
The parallel-reduction Rayon result ($+$192.2\,\%, CV\,12\,\%) is the most
anomalous: the high CV indicates measurement instability, and the result
may reflect suboptimal Rayon configuration rather than an inherent Rust
limitation.

\subsection{GPU Benchmark Outcomes}
\label{sec:gpu_results}

Table~\ref{tab:gpu_results} presents the GPU kernel results with full
statistical detail.
Timing covers kernel execution only; H2D and D2H transfer times are
excluded.
Note that all three GPU benchmarks exhibit high CV\,\% values due to the
small trial count ($n = 5$), and results should be interpreted with
appropriate caution.

\begin{table}[htbp]
\caption{GPU kernel performance: C++ vs.\ Rust (kernel time only,
         H2D/D2H transfer excluded, $n=5$ trials per benchmark).}
\label{tab:gpu_results}
\centering
\begin{tabular}{llrrrrrr}
\toprule
Benchmark & Size & C++ (ms) & Rust (ms) & Overhead & $p$-value &
Cohen's $d$ & CV\,\% (C++/Rust) \\
\midrule
cuda\_matmul\_tiled & 1024     & 0.0018 $\pm$ 0.0009 & 0.0002 $\pm$ 0.0001 & $-$86.5\,\% & 0.017$^{*}$  & $-$2.49 & 49.0 / 22.8 \\
cuda\_reduction     & 1\,000\,000 & 0.0633 $\pm$ 0.0162 & 0.0002 $\pm$ 0.0001 & $-$99.6\,\% & 0.001$^{***}$ & $-$5.51 & 25.6 / 22.8 \\
cuda\_softmax       & 64$\times$16384 & 0.0012 $\pm$ 0.0012 & 0.0017 $\pm$ 0.0033 & $+$37.7\,\% & 0.778 (ns)   & $+$0.19 & 96.2 / 193.7 \\
\bottomrule
\end{tabular}
{\small $^{*}p<0.05$, $^{***}p<0.001$, ns\,=\,not significant.
 Overhead positive means Rust is slower.
 CV\,\% = coefficient of variation; values $>$5\,\% indicate high
 measurement instability.}
\end{table}

The matrix multiplication and reduction kernels both yield Rust advantages
of large to very large effect size ($|d| > 2.49$).
However, the C++ matmul CV\,\% of 49\,\% indicates substantial between-trial
variability, which weakens confidence in the point estimates despite the
statistically significant $p$-value.
The reduction result is more stable (C++ CV\,25.6\,\%, Rust CV\,22.8\,\%)
and aligns with the expected behaviour of a memory-bandwidth-bound kernel
where driver overhead---absorbed differently in the C++ and Rust harnesses---
may account for part of the difference.
The softmax comparison ($p = 0.778$, $d = 0.19$) provides no evidence of a
difference between the two implementations; both exhibit very high CV\,\%
values (96\,\% and 194\,\% respectively), which may reflect warm-up
variability in first-call kernel launch latency with only five trials.
A larger trial count would be needed before any conclusion could be drawn
for the softmax benchmark.

\subsection{Safety and Code Metrics}
\label{sec:safety}

Table~\ref{tab:safety} reproduces the safety-oriented metadata from the
committed \texttt{results/raw/safety\_metrics.json} file.

\begin{table}[htbp]
\caption{Safety and code-size metrics from \texttt{safety\_metrics.json}.}
\label{tab:safety}
\centering
\begin{tabular}{lr}
\toprule
Metric & Value \\
\midrule
C++ CPU LOC & 739 \\
C++ GPU LOC & 709 \\
Rust CPU LOC & 731 \\
Rust GPU LOC & 141 \\
Aggregate Rust/C++ LOC ratio & 0.60 \\
Recorded Rust \texttt{unsafe} blocks (measured paths) & 0 \\
\bottomrule
\end{tabular}
{\small LOC = non-blank, non-comment lines in \texttt{cpu/} and
 \texttt{gpu/} source trees; \texttt{build/} and \texttt{target/}
 excluded.
 ``Measured paths'' refers to \texttt{cpu/rust/} and
 \texttt{gpu/rust/} only; transitive \texttt{unsafe} in
 Cargo dependencies is not counted.}
\end{table}

The Rust GPU code is substantially shorter (141 vs.\ 709 lines), which is
partly explained by the hybrid architecture: the Rust GPU runner relies on
precompiled CUDA PTX kernels loaded at runtime rather than embedding full
kernel source.
The zero \texttt{unsafe} count indicates that none of the measured source
paths---\texttt{cpu/rust/} and \texttt{gpu/rust/}---contain the
\texttt{unsafe} keyword.
This is consistent with H3, but as noted in Section~\ref{sec:methodology},
it does not account for \texttt{unsafe} code inside Cargo dependencies
or the PTX-compiled kernel code.
LOC and \texttt{unsafe} counts are only partial proxies for safety and
maintainability; they should be treated as supporting evidence rather than
conclusive proof that Rust implementations are objectively safer or easier
to maintain.

% ─────────────────────────────────────────────────────────────
\section{Discussion}

\subsection{Interpretation of Results}

The CPU results are broadly consistent with Ivanov's finding that Rust can
match or exceed C++ on optimised standard-library routines~\cite{ivanov2022rust}.
The search, sorting, and hash-table advantages likely reflect stronger
algorithmic choices in the Rust standard library (e.g., pdqsort) rather than
a pure language overhead difference---a nuance that underscores the
importance of distinguishing ``language performance'' from ``standard-library
performance.''
The FFT and naive matrix multiplication disadvantages are consistent with
the observation in Martins~et~al.\ that Rust's auto-vectorisation can lag
behind GCC on certain numerical kernels~\cite{martins2025npb}; this is an
active area of improvement in the LLVM back-end shared by both
\texttt{rustc} and Clang.

The GPU results are more difficult to interpret with confidence.
The matmul and reduction outcomes favour Rust by large margins, but both
benchmarks show high CV\,\%, and the small trial count ($n=5$) limits
statistical power.
These results are directionally consistent with Bychkov and Nikolskiy's
observation that Rust GPU drivers can match C++ when kernel code is
equivalent~\cite{bychkov2022rust}, but the variability means the magnitude
of the advantage is uncertain.
The softmax non-result ($p = 0.778$) is the most honest finding in the GPU
section: with only five trials and CV values exceeding 96\,\%, no conclusion
about relative performance is warranted.

The zero \texttt{unsafe} count across measured Rust paths supports H3 and
is consistent with the broader claim in Jung~et~al.\ that the Rust type
system can enforce safety without runtime cost~\cite{jung2017rustbelt}.
However, as Fulton~et~al.\ document, safety benefits can be partially offset
by increased development effort~\cite{fulton2021benefits}; LOC alone does
not capture that trade-off.

\subsection{Implications}

For practitioners deciding whether to adopt Rust for a new high-performance
system, these results suggest that Rust is a credible choice for workloads
similar to search, sorting, and reduction, and that GPU kernel execution
performance is not inherently disadvantaged by choosing Rust tooling.
The FFT and matrix-multiplication gaps suggest caution for workloads that
depend heavily on numerically mature C++ libraries or SIMD optimisations
that Rust's LLVM back-end has not yet fully exploited.

\subsection{Limitations and Threats to Validity}

Several threats limit the generalisability of these results.
The benchmark suite is small and selective; results may not extend to
workloads such as graph algorithms, sparse linear algebra, or distributed
memory communication.
All results are tied to one hardware configuration; the Core Ultra~7's
hybrid P/E-core topology and the RTX~5060~Ti's Blackwell memory subsystem
may produce different relative results than server-class hardware.
The GPU trial count ($n=5$) is insufficient for reliable statistical
inference on high-variance kernels; the softmax result cannot be
interpreted.
Finally, LOC and \texttt{unsafe} counts are imperfect proxies.
They do not measure compile-time complexity, dependency safety, or the
cognitive overhead of the borrow checker.

\subsection{Suggestions for Future Research}

Future work should increase GPU trial counts to at least 30, add warm-up
iterations to reduce first-call latency, and extend the benchmark suite to
include sparse and irregular workloads.
A profiling study examining auto-vectorisation decisions for the FFT and
matmul benchmarks would clarify whether the observed C++ advantage is
architectural or compiler-strategy dependent.
Measuring compile times (currently \texttt{null} in the repository
metadata due to toolchain availability constraints) would also strengthen
the H3 analysis.

% ─────────────────────────────────────────────────────────────
\section{Conclusion}

\subsection{Summary of Main Findings}

This study compared Rust and C++ on 15 CPU benchmarks and 3 GPU kernels
using a controlled hardware environment, identical input datasets, and
rigorous statistical analysis.
Rust outperforms C++ on search, sorting, and hash-table workloads by
statistically significant margins.
C++ retains advantages on FFT, naive matrix multiplication, and one
parallel-reduction configuration.
GPU kernel comparisons favour Rust for matrix multiplication and reduction,
though high measurement variance constrains confidence in those results,
and the softmax benchmark shows no statistically significant difference.

\subsection{Research Contribution}

This work contributes a unified, reproducible benchmark artifact for
Rust-vs-C++ evaluation spanning both CPU and GPU domains---an evaluation
that prior literature has not provided in a single, statistically rigorous
study.
The publicly committed raw data and analysis scripts allow independent
verification and extension.

\subsection{Closing Thoughts}

The evidence from this controlled study supports the view that Rust is a
viable language for performance-sensitive CPU and GPU programming.
It is not universally faster than C++, and workload characteristics
matter substantially.
However, the results demonstrate that competitive performance can be achieved
without resorting to \texttt{unsafe} code in the measured implementations,
which suggests that Rust's safety model and high performance are not
mutually exclusive goals---at least for the class of workloads examined here.

\medskip
\printbibliography

\end{document}
